Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: minor changes.

In this section, we discuss some of the software projects as case
studies to better understand the points discussed during the workshop
and described in the previous sections. We dive deeper to find how are
they affected by the sustainability related issues in
practice. Additionally, we discuss issues surrounding people,
especially the career paths of original developers and investigate to
answer the question: \emph{are the original developers still around?}

We classify the software projects discussed in the workshop in two
broad categories. First, the \emph{utility} software, comprising of
general purpose software. Utility software is often used as enabler
and/or facilitator for the development of other tools and techniques
to carry out scientific work. This includes the software developed to
efficiently utilize new research infrastructures. Second, the
\emph{scientific} software, comprising the software that was
originally developed with an aim to solve a specific scientific
problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different.  Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or visualization software products such as
\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
\textit{ab initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}.

@@

% What are the sustainability challenges? What are the opportunities?
Development is often a high risk/reward undertaking exploring
uncharted territories and is largely influenced by (re)usability
factors. Owing to relatively large number of features, the development
and prototype process is also lengthy which poses a significant
survival risk. Challenges on a class of utility software for new
architectures is well discussed in~\cite{Ferenbaugh_WSSSPE}.

On the other hand, utility software presents opportunities to be
usable by a larger community making its undertaking and development an
attractive pursuit.  It is generally more visible in community which
in turn leads to a broader and deeper participation. For instance,
promoting collaborations across the breadth (e.g., different
departments) and depth (e.g., stakeholders within a department) of
community, one of the key ingredient of a sustainable
process. Successful utility projects reap high rewards and have a
longer usage span. Development process becomes user-driven and
self-sustaining.

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
follows agile software development practices and implements standards
such as test-driven development and socialized bug managing practices
via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
offers easy community sharing of data and tools further promoting a
collaborative environment. The project very closely follows the
guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
software projects are often developed aiming better utilizing a
particular, new infrastructure and architecture, e.g.,
\toolname{MVAPICH}, \toolname{VisIT}, toolname{ParaView}. Similarly, to leverage the power
of emerging architectures such as accelerators, new code and libraries
are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
but nonetheless with many invaluable lessons were learned about
influential cultural and technical aspects in sustainable software
development practices.

@@

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software
that was purpose-built for a target use-case scenario, fixed/frozen
requirements in mind or solving a specific problem. Software projects
pertaining to specific scientific domain tend to be in a niche and the
user community tends to be small to medium. They are mostly driven by
the science and specific needs of a research group. Specific needs
such as numerical accuracy and algorithmic optimization are some of
the paramount requirements of most scientific software.

% What are the challenges?
Long-term sustainability of scientific software is often a significant
challenge and face radically different issues compared to utility
software.  Many submissions reported that software is practically
considered a ``byproduct'' of the actual research. Others contended
that the software was not the main funded part of their research. A
smaller codebase and fixed requirements result in stability, ease of
installation, and configurability.  Many such projects mature and are
treated as libraries to be used into larger systems such as some of
the utility software discussed in the previous section. While the
software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a very few developers who
might not be specialists in software development. Development tends to
be linear and simplistic with a limited scope to follow software best
practices.

Some examples of such software discussed as part of the workshop are,
\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware},
\toolname{PETSc}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
the above efforts are focused on one scientific and/or applied
mathematics domain. However, sometimes such projects grow beyond the
initial vision of developers. One such example is \toolname{Kitware},
which while being a software product specializing in the scientific
process, has a core focus of developing communities around software
processes. An instance of this process is the development of the
\toolname{CMake} build utility, which started out as a building tool
for \toolname{ITK} but grew to become a generic build utility for C++
projects. Similarly, \toolname{PETSc} is growing towards becoming a
general purpose utility system usable for solving variety of
scientific problems.

Finally, it appears that the career options of developers from
different kinds of software projects differ significantly. In order to
further investigate the much discussed issue of career paths for
developers of these software, we conducted a small survey of the cases
discussed above with respect to the current status of its core
developers. We looked at the workshop submissions and the original
cited papers for the respective software. Where available, we looked
at the web pages and source code of the earliest versions to find the
original developers. We included projects older than 5 years and
excluded relatively new projects such as \toolname{SciPy} and
\toolname{numFOCUS} from this survey. We find that in a majority of
scientific software, more than one original developer are still
around, in some cases after more than 10 years (e.g.,
\toolname{R/qtl}).  Whereas, in the case of utility software projects,
it appears that at least one original developer is still around, often
in a part-time or ``benefactor'' role. In other cases, it appears that
the original developers have went on to assume a larger role, with the
same project.

In conclusion, we find that there are distinctions in the
characteristics and the kinds of challenges faced between utility and
scientific software projects.  In terms of career paths of original
developers, it appears that the career choices for utility software
developers are broader compared to the developers of scientific
software. However, it appears that in the case of scientific software,
core competencies of original developers tend to be domain science who
happen to be the coders of their software. As projects as a whole, the
small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. We also find that the
utility software tend to adhere more loyally to the many best
practices discussed during the workshop. Often the sustainability of
scientific software projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science,
e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
scientific software requires more scientific skills compared with
those of utility software. This diminishes the need of specialist
programmers and practices in many cases where the bulk of development
is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and/or principles for sustainability
be defined and practiced for each category of scientific software and
they be given separate treatment.

>>>>

In this section, we discuss some of the software projects as case
studies to better understand the points discussed during the workshop
and described in the previous sections. We dive deeper to find how 
they are affected by the sustainability related issues in
practice. Additionally, we discuss issues surrounding people,
especially the career paths of original developers and investigate to
answer the question: \emph{are the original developers still around?}

We classify the software projects discussed in the workshop in two
broad categories. First, the \emph{utility} software, comprising of
general purpose software. Utility software is often used as enabler
or facilitator for the development of other tools and techniques
to carry out scientific work. This includes the software developed to
efficiently utilize new research infrastructures. Second, the
\emph{scientific} software, comprising the software that was
originally developed with an aim to solve a specific scientific
problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different. Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or visualization software pinroducts such as
\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
\textit{ab initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}. \todo{cite}

@@

% What are the sustainability challenges? What are the opportunities?
Development is often a high risk/reward undertaking exploring
uncharted territories and is largely influenced by (re)usability
factors. Owing to a relatively large number of features, the development
and prototype process is also lengthy which poses a significant
survival risk. Challenges on a class of utility software for new
architectures is well discussed in~\cite{Ferenbaugh_WSSSPE}.

On the other hand, utility software presents opportunities to be
usable by a larger community making its undertaking and development an
attractive pursuit.  It is generally more visible in community which
in turn leads to a broader and deeper participation. For instance,
promoting collaborations across the breadth (e.g., different
departments) and depth (e.g., stakeholders within a department) of
community, one of the key ingredients of a sustainable
process. Successful utility projects reap high rewards and have a
longer usage span. Development process becomes user-driven and
self-sustaining.

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
follows agile software development practices and implements standards
such as test-driven development and socialized bug managing practices
via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
offers easy community sharing of data and tools further promoting a
collaborative environment. The project closely follows the
guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
software projects are often developed aiming better utilizing a
particular, new infrastructure and architecture, e.g.,
\toolname{MVAPICH}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to leverage the power
of emerging architectures such as accelerators, new code and libraries
are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
but nonetheless with many invaluable lessons were learned about
influential cultural and technical aspects in sustainable software
development practices.

@@

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software
that was purpose-built for a target use-case scenario, fixed 
requirements in mind, or solving a specific problem. Software projects
pertaining to specific scientific domain tend to be in a niche and the
user community tends to be small to medium. They are mostly driven by
the science and specific needs of a research group. Specific needs
such as numerical accuracy and algorithmic optimization are some of
the paramount requirements of most scientific software.

% What are the challenges?
Long-term sustainability of scientific software is often a significant
challenge and face radically different issues compared to utility
software.  Many submissions reported that software is practically
considered a ``byproduct'' of the actual research. Others contended
that the software was not the main funded part of their research. A
smaller codebase and fixed requirements result in stability, ease of
installation, and configurability.  Many such projects mature and are
treated as libraries to be integrated into larger systems such as some of
the utility software discussed in the previous section. While the
software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a few developers who
are often not specialists in software development. Development tends to
be linear and simplistic with a limited scope to follow software best
practices.

Some examples of such software discussed as part of the workshop are,
\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware},
\toolname{PETSc}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
the above efforts are focused on one scientific or applied
mathematics domain. However, sometimes such projects grow beyond the
initial vision of developers. One such example is \toolname{Kitware},
which while being a software product specializing in the scientific
process, has a core focus of developing communities around software
processes. Another instance of this process is the development of the
\toolname{CMake} build utility, which started out as a building tool
for \toolname{ITK} but grew to become a generic build utility for C++
projects. Similarly, \toolname{PETSc} is growing towards becoming a
general purpose utility system usable for solving a variety of
scientific problems.

Finally, it appears that the career options of developers from
two kinds of software projects differ significantly. In order to
further investigate the much discussed issue of career paths for
developers of these software, we conducted a small survey of the cases
discussed above with respect to the current status of its core
developers. We looked at the workshop submissions and the original
cited papers for the respective software. Where available, we looked
at the web pages and source code of the earliest versions to find the
original developers. We included projects older than 5 years and
excluded relatively new projects such as \toolname{SciPy} and
\toolname{numFOCUS} from this survey. We find that in a majority of
scientific software, more than one original developer are still
around, in some cases after more than 10 years (e.g.,
\toolname{R/qtl}).  Whereas, in the case of utility software projects,
it appears that at least one original developer is still around, often
in a part-time or ``benefactor'' role. In other cases, it appears that
the original developers have went on to assume a larger role with the
same project.

In conclusion, we find that there are distinctions in the
characteristics and the kinds of challenges faced between utility and
scientific software projects.  In terms of career paths of original
developers, it appears that the career choices for utility software
developers are broader compared to the developers of scientific
software. However, it appears that in the case of scientific software,
core competencies of original developers tend to be domain science who
happen to be the coders of their software. As projects as a whole, the
small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. We also find that the
utility software tend to adhere more loyally to the many best
practices discussed during the workshop. Often the sustainability of
scientific software projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science,
e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
scientific software requires more scientific skills compared with
those of utility software. This diminishes the need of specialist
programmers and practices in many cases where the bulk of development
is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and principles for sustainability are to
be defined and practiced for each category of scientific software and
they be given separate treatment.
